---
title: "Exploratory Analysis of Conflict Dataset"
author: "Mohsyn Imran Malik"
format: pdf
editor: visual
---

## Exploratory Analysis

Here is our final dataset created from last week

`combo_df`

First, we will load the `tidyverse and dplyr` package to aid in our exploratory analysis

```{r}
library(tidyverse)
library(dplyr)
```

Lets start by looking at head and tail of dataset

```{r}
head(combo_df)

tail(combo_df)
```

We can print summary stats (mean, median and interquartile range) of our dataset and get an idea of the missing data (NA)

```{r}
summary_df <- summary(combo_df) 

summary(combo_df)
```

We can also summarize by country or year (warning: large data output)

```{r}
summary_country <- by(combo_df, combo_df$ISO, summary)
summary_country

summary_year <- by(combo_df, combo_df$Year, summary) 
summary_year
```

Note that we have significant amounts of missing data (NA) for a number of variables in the dataset.

Lets compare mean and standard deviation of continuous variables data when we remove all rows with missing data

```{r}
c <- c(3:5, 9, 12, 15:17, 19:22) 

mean <- sapply(combo_df[, c], mean, na.rm = TRUE) 

mean 

sd <- sapply(combo_df[, c], sd, na.rm = TRUE) 

sd
```

Now lets remove rows with missing data and find new means and sd

```{r}
remove_na <- combo_df[complete.cases(combo_df[ , c]),] 

rem_mean <- sapply(remove_na[, c], mean, na.rm = TRUE) 

rem_mean

 rem_sd <- sapply(remove_na[, c], sd, na.rm = TRUE) 

rem_sd
```

We can do the same but this time by using multiple imputations to account for missing data

```{r}
library(mice)
mult_imp <- mice(combo_df, print = FALSE) 
mice_est <- complete(mult_imp) |> as_tibble() 

imp_mean <- sapply(mice_est[, c], mean, na.rm = TRUE) 
imp_mean 
imp_sd <- sapply(mice_est[, c], sd, na.rm = TRUE) 
imp_sd
```

Use cbind to visualize the differences in means on method of dealing with missing data in table format

```{r}
md_df <- cbind(mean, rem_mean, imp_mean, sd, rem_sd, imp_sd) 

md_df<- as.data.frame(md_df) 

md_df<- rownames_to_column(md_df, var = "variables") 

md_df
```

Lets make a visual plot of contingency tables using mosaicplots.\
We can look at year and binary conflict variable.

```{r}
 mosaicplot(table(combo_df$Year, combo_df$bin_conflict), color = TRUE, xlab = "Year", ylab = "Presence of Conflict in country")
```

We can visualize the spread of data using histograms for our continuous variables #Lets look at total droughts, total earthquakes, total battle related death, gdp100, popdens, male_edu, temp, rainfall1000, maternal.mortality, infant.mortality, neonatal.mortality, and under 5 mortality

```{r}
num_combo_df <- sapply(combo_df, as.numeric) 

for (i in c(3:5, 9, 12, 15:17, 19:22)) { hist(num_combo_df[, i], main = paste("Histogram of", names(combo_df[,i])), col = "lightblue") }
```

Lets use a bar graft to examine the binary variable of conflict for all countries and years included

```{r}
ggplot(data = combo_df) + geom_bar(mapping = aes(x = bin_conflict))
```

Lets plot gdp100 by conflict using ggplot

```{r}
ggplot(data = combo_df) + geom_point(mapping = aes(x = gdp1000, y = bin_conflict))
```

Lets make a comparative histogram to see how bins of gdp compare in those with conflict and those without.

```{r}
 ggplot(data = combo_df, aes(x = gdp1000, group = bin_conflict, fill = bin_conflict )) + geom_histogram(position='dodge')+theme_classic()
```

Lets further examine this relationship by creating a new variable which divides data into quartiles based on male education

```{r}
combo_df$male_edu_qt <- ntile(combo_df$male_edu, 4) 
ggplot(data = combo_df, aes(x = gdp1000, group = bin_conflict, fill = bin_conflict )) + geom_histogram(position='dodge')+theme_classic()+facet_wrap(vars(male_edu_qt))
```

Let us examine the trend in maternal mortality over the years and stratify into separate plots based on conflict

```{r}
ggplot(data = combo_df, aes(x = Year, y = Maternal.Mortality)) + geom_point() + geom_smooth() + facet_wrap(vars(bin_conflict))
```

Lets do the same plot, but use our dataset with multiply imputed missing variables

```{r}
ggplot(data = mice_est, aes(x = Year, y = Maternal.Mortality)) + geom_point() + geom_smooth() + facet_wrap(vars(bin_conflict))
```
